MSstatsPTM: Statistical relative quantification of post-translational modifications in bottom-up mass spectrometry-based proteomics

### 2022-07-07 main/MSstatsPTM_Main.tex
Updated methods sections

\medskip \noindent 
\todo{OV: this doesn't look right, as it seems that all the features and replicates equally contribute to the error. Do people do some kind of summarization of the features first? Let's first describe the summarization}
\todo{DK: I think we should just delete this part entirely. It isn't how we compared the method and is not really used in the literature from what I can tell.} \todo{OV: below in the results we use the notation $y_{i++}$ etc. Alternatively, since we put a hat on $\hat{y}_{ij}$, maybe we do not need '++', as it is clear that this is a summary. I removed '++' in the results section for now. Let's use the same notation in this section consistently}

@@

\medskip \noindent 
The basic $t$-test can be extended to account for the confounding of changes in PTM abundance and overall changes in protein abundance~\cite{Schwammle2015,THOMAS2020,Mertins:2013}. Typically this is done by first summarizing the feature intensities in each run, separately for modified and unmodified features, e.g. with the sum on the original scale, dividing the summary of the PTM abundance by the summary of the protein abundance and then taking a log. Specifically, for condition $i$ and biological replicate $j$, the approach calculates
\begin{eqnarray}
u_{ij} = \log \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right) - \log \left( \sum_{l=1}^{L} 2^{y_{ijl}} \right)
\label{eq:adj_estimation}
\end{eqnarray}
The approach then models these values as
\begin{eqnarray}
u_{ij} = (\mu^{\ast}_{i}-\mu_i) + \epsilon^{\prime}_{ij}, \text{where } \epsilon^{\prime}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{\prime 2}), i \in \{1, 2\}
\label{eq:ttest_model}
\end{eqnarray}
The model allows us to estimate $\hat{\Delta}_{adj}$ and its standard error.
Based on this model, we can test the more relevant null hypothesis in~\eqref{null_hyp}, by comparing the test statistic against the Student distribution with \todo{how many} degrees of freedom in balanced designs.

Although effective, the approach is somewhat simplistic. It is not applicable to experimental designs with more complex sources of biological and technological variation, such as experiments with repeated measurements, experiments with multiple batches or experiments with multi-run multiplexing. Since \eqref{adj_estimation} performs the adjustment on the replicate level, the experiment must contain a matching number of replicates in both the modified and unmodified runs. Technological artifacts such as missing values further undermine the calculation of $u_{ij}$ in~\eqref{adj_estimation}. Finally, the lack of a self contained, straightforward implementation of the method challenges the applications \todo{OV: is the last statement true? Seems simple enough}. \todo{DK: I meant more that there was no simple package implementation. Researchers need to program the methods themselves. I cant find an implementation at least}

@@

$MSstats$~\cite{Choi:2014} and $MSstatsTMT$~\cite{Huang:2020} are a series of R/Bioconductor packages designed for protein significance analysis and statistical relative quantification of proteins and peptides in global, targeted and data-independent proteomics. Models in these packages takes as input a list of log$_2$-intensities $y_{ijk}$. For each protein, the feature log$_2$-intensities are first summarized into a single value per protein per run using Tukey's median polish~\cite{Tukey:1977}. These summarized values are then used as input to fit a flexible family of linear mixed-effects models \cite{McLean:1991, Faraway:2006, Bolker2009}. The specific model instances depend on the experimental design and data acquisition. For example, for an experiment with \todo{group comparison or time course design?} $MSstatsTMT$ fits the following model \todo{OV: could we keep the notation as consistent as possible across this section? Above conditions had subscripts $i$ and bio reps had subscripts $j$. Could we keep a consistent notation below (or change the notation elsewhere in the manuscript?)} \todo{Should we maybe give a model for label-free, to stay consistent with the example in figure 2? Just use $I$ and $J$ etc for the indices...}
\begin{eqnarray}
& y_{mtcb} = \mu + Mixture_m + TechRep(Mixture)_{t(m)} + Condition_c + Subject_{mcb} + \epsilon_{mtcb}, & \\ 
& \text{where }  Mixture_m  \mathop\sim^{iid} \mathcal{N}(0, \sigma^2_M),  TechRep(Mixture) \mathop\sim^{iid} \mathcal{N}(0, \sigma^2_T), &  \nonumber \\
& \sum\limits_{c=1}^{C} Condition_c = 0,\ \todo{Subject \stackrel{iid}{\sim}\mathcal{N}(0,?)}\  \epsilon_{mtcb} \mathop\sim^{iid} \mathcal{N}(0, \sigma^2) & \nonumber
\end{eqnarray}
Parameters of the model are estimated using restricted maximum likelihood (REML)~\cite{Kenward}.
In turn, the parameters allow us to estimate the relevant quantity $\hat{\Delta}_{Protein}$ \todo{this has not been defined, Define here, using the notation above?} and its standard error.
Similarly to $Limma$, $MSstatsTMT$ includes an optional Empirical Bayes moderation of the standard error~\cite{Huang:2020}. 
\todo{OV: maybe elaborate on that?} Based on the model, the approach tests the null hypothesis in \eqref{conv_null_hyp}, by comparing the test statistic against the Student distribution.

The $MSstats$ framework has a number advantages over the methods above. 
First, unlike \todo{t-test (or ANOVA?) and $Limma$}, the family of linear mixed effects models is applicable to many complex experimental designs, including designs with multiple sources of variation, and unbalanced designs. 
Second, the approach is applicable to various data acquisition types, including label-free DDA and DIA, and experiments with TMT labeling, and is compatible with various data processing tools such as \todo{XXX} 
Third, the run-level summarization has been shown to be more robust to outliers and missing values as compared to feature averages~\cite{Choi:2014}. 
Finally, the custom $MSstats$ and $MSstatsTMT$ implementation accounts for potential data artifacts, and is numerically scalable and stable, and \todo{is available through both command line and a dedicated graphical user interface}.
Unfortunately, MSstats estimates the difference in overall protein abundance and tests the null hypothesis in~\eqref{conv_null_hyp}. It does not account for the confounding between the changes in PTM abundance and the overall changes in protein abundance. In other words, it cannot directly test the null hypothesis in~\eqref{null_hyp}.

@@

We leverage the existing methods in $MSstats$ and $MSstatsTMT$ and extend them to remove confounding between changes in the PTM and the overall changes in protein abundance.  Each feature type is summarized separately using $MSstats$ or $MSstatsTMT$, resulting in run level values for the PTM, \todo{$\hat{y}_{ij}^{\ast}$} \todo{OV: I am confused: since we summarize per run, we need at least an index for a condition and an index for bio rep. Could we use this consistently across all the sections?}, and unmodified peptide. The estimated summaries of the modified features are then used as the input to the model \ov{such as in eq XXX} in $MSstats$ or $MSstatsTMT$. Similarly, the estimated summaries of the unmodified features are used as the input to the corresponding model in $MSstats$ or $MSstatsTMT$.

For the modified features, the model-based estimates in $MSstats$ and $MSstatsTMT$ include $\hat{\Delta}_{PTM}=\hat{\mu}^{\ast}_{i}-\hat{\mu}_{i^{\prime}}^{\ast}$, and its standard error $\widehat{SE}(\hat{\Delta}_{PTM})$. Similarly, for the unmodified features, the model-based estimates in $MSstats$ and $MSstatsTMT$ incluse $\hat{\Delta}_{protein}=\hat{\mu}_{i}-\hat{\mu}_{i^{\prime}}$, and its standard error $\widehat{SE}(\hat{\Delta}_{protein})$. Therefore, the adjusted difference $\hat{\Delta}_{adj}$ in~\eqref{null_hyp} can be estimated as
\begin{eqnarray}
\hat{\Delta}_{adj} = (\hat{\mu}^{\ast}_{i}-\hat{\mu}_{i}) - (\hat{\mu}_{i^{\prime}}^{\ast}-\hat{\mu}_{i^{\prime}}) = (\hat{\mu}^{\ast}_{i}-\hat{\mu}_{i^{\prime}}^{\ast}) - (\hat{\mu}_{i}-\hat{\mu}_{i^{\prime}}) = \hat{\Delta}_{PTM} - \hat{\Delta}_{protein} 
\label{eq:Delta_hat_adj}
\end{eqnarray}
Assuming that the unexplained by the model  sources of variation in the modified features are independent from the corresponding sources of variation in the modified features, the standard error $\widehat{SE}(\hat{\Delta}_{adj})$ is obtained by combining the standard errors from the unmodified and modified model fits.
\begin{eqnarray}
\widehat{SE}(\hat{\Delta})_{adj} = \sqrt{ \widehat{SE}(\hat{\Delta}_{PTM})^{2} + \widehat{SE}(\hat{\Delta}_{protein})^{2}} 
\label{eq:prop_se}
\end{eqnarray}
The estimated standard error is larger than the standard errors associated with each distinct feature type, reflecting the increased uncertainty inherent to combining the uncertainty in two estimators. Finally, the degrees of freedom associated with~\eqref{prop_se} are obtained via the Satterthwaite approximation~\cite{satterthwaite:1946}\todo{OV: maybe site a more recent exp design book?}
\begin{eqnarray}
\left( \widehat{SE}(\hat{\Delta}_{PTM})^{2} + \widehat{SE}(\hat{\Delta}_{protein})^{2} \right)^2 \bigg/
\left( \frac{\widehat{SE}(\hat{\Delta}_{PTM})^{4}}{\mathrm{df}(\widehat{SE}(\hat{\Delta}_{PTM}))} + \frac{\widehat{SE}(\hat{\Delta}_{protein})^{4}}{ \mathrm{df}(\widehat{SE}(\hat{\Delta}_{protein}))} \right)
\label{eq:prop_df}
\end{eqnarray}
To test the null hypothesis in~\eqref{null_hyp}, the test statistic $\hat{\Delta}_{adj}/\widehat{SE}(\Delta)_{adj}$ is compared with the Student distribution with the degrees of freedom in~\eqref{prop_df}. The p-values of the comparison are adjusted for multiple testing using the approach by Benjamin and Hochberg~\cite{Benjamini:1995}.

By leveraging the implementations in $MSstats$ and $MSstatsTMT$, the proposed approach is very versatile. It is applicable to a wide variety of experimental designs, including group comparison, paired designs, time course designs and unbalanced designs. It is applicable to label-free data acquisition strategies such as DDA and DIA and experiments with TMT labeling. It is compatible with data processing tools \todo{XXX}. The proposed approach is also applicable to experiments where the experimental designs for PTM profiling and global proteome profiling vary in properties such as number of biological replicates, data acquisition strategies and runs. \todo{OV: We do have a requirement that the scale of measurements is comparable between TMT and PTM, correct? Otherwise the SE may not be comparable}

@@

\todo{OV: I feel that some parts of this section are either specific to MSstats/MSstatsTMT, or a bit redundant with the methods. For the parts that we do not implement anew but reuse, could you move them to the 'Existing methods'/MSstats section? Here let us focus only on things that we had to develop from scratch for this project. If we implement the methods, we do not need to re-describe the methods}

\todo{OV: could you make this section more specific to PTM converters? E.g., I remember that the input should be different, there are some parts relevant to the localization of PTMs. This was not part of the base MSstats. Could you emphasize that?} The proposed methods are implemented in the open source R package $MSstatsPTM$. An overview of the steps of the package are illustrated in Figure \ref{fig:msstatsptm_design}. $MSstatsPTM$ includes converters for multiple spectral processing tools, including MaxQuant, Progenesis, and Spectronaut. The converters take as input the raw data from the tool, identify the modification site for modified peptide, and put the data into the correct format for analysis in $MSstatsPTM$. Conversion is done separately for the modified peptide and global profiling runs. If the global profiling information is not available, the package can still analyze the modified run, but will do so without adjusting for changes in unmodified protein abundance. This amounts to testing the null hypothesis in~\eqref{conv_null_hyp}; which amounts to running the methods seen in $MSstats$ on modified peptide level data \cite{Choi:2014} \cite{Huang:2020}.

\todo{OV: this is the part that looks redundant with the MSstats section. In the MSstats section, do not repeat the methods, just focus on the specifics of the implementation. Relavant parts would be that we fit simplified models if the base model is not applicable. In this section,  could you just say that we do the same as MSstats (including fitting simplified models). But there are some caveats, e.g. assumptions of missing values due to low abundance are not necessarily appropriate here so we recommend to not use that option? Also specialized sample size calculations} After using the converters, the next step is peptide/protein summarization and missing value imputation. For the modified run, the package summarizes features which include the same modification together. Features with multiple modification sites are not included with single site features and are summarized separately. For the global profiling run all unmodified features from the same protein are summarized together, up to the protein level. Additionally, the summarization includes global median normalization and normalizes between MS runs. The package can optionally impute missing values using an AFT model~\cite{Wei:1992}. Missing value imputation should be performed with care when applied to PTM data. PTMs generally exhibit low feature counts and may be missing due to reasons other than low abundance. These issues can cause the missing value imputation to be unstable.

The final step of the package is modeling the summarized dataset. A linear mixed effects model is fit for both the summarized modified and global profiling runs. This model is automatically adjusted depending on the experimental design and acquisition method. The comparisons of interest can either be predefined or a full pairwise comparison will be tested. After fitting a model to both the modified and unmodified data, the modified model is adjusted for changes in unmodified protein abundance, using the methods described above.

\todo{OV: similarly, some of this belongs to MSstats. Could you focus on project-specific implementation. E.g., we separately visualize modified and unmodified peptides. (Btw, sequence-specific visualization such as the one you did for MSstatsLiP would be awesome if there is a chance)} Beyond the core functionality, the package also includes functions for plotting the results. These include summarization plots to help with quality assurance and identifying sources of variation (such as quality control plots and profile plots), as well as modeling plots (such as volcano plots and heatmaps), which evaluate the fold change between conditions and peptides.

@@

The evaluations were based on the $MSstatsPTM$ \todo{any details of the options? E.g., was Empirical Bayes moderation used in MSstatsPTM?} $MSstatsPTM$ was compared \todo{with methods XXX - why chosen? We did not compare to IsobarPTM because XXX. Do we say t-test or ANOVA?} All the evaluations were done at the adjusted p-value cutoff of $\alpha=.05$. More details are in {\bf Supplementary Sec. \ref{supp-sec:experiments}}.

@@

\figref{sim1_fdr} illustrated the high FDR observed on the simulated datasets when not accounting for the overall changes in protein abundance. After the adjustment, the FDR of all methods improved. In a clean simulation, recall and accuracy of $MSstatsPTM$ and $Limma$ performed similarly. \todo{As expected when not using EB moderation in MSstatsPTM?,} Limma performed slightly better in simulations with fewer biological replicates. In contrast, the performance of the two-sample $t$-test method lagged behind. This is due to the fact that the $t$-test only uses data within the groups of interest while ignoring the remainder of the data. In contrast, $MSstatsPTM$ and Limma leverage information across comparisons, resulting in improved power in simulations with more comparisons.

@@

\todo{This section just describes the figure. Could you emphasize why we see these results? What are the reasons for differences in performance?} In simulations with missing values and few features, $MSstatsPTM$ outperformed $t$-test and $Limma$. Changes in unmodified protein level still needed to be accounted for to control the FDR. Once controlling for changes in the unmodified protein, the proposed method outperformed the other methods, as seen in~\figref{sim2_acc}. The proposed method calibrated model accuracy and recall rate well, even when the number of replicates were low. Additionally, when comparing the fold change estimation across all modified peptides, the proposed method showed a tighter distribution of estimated fold changes around the true fold change, as compared to $t$-test and $Limma$. Specifically, the inner quartile range (IQR) of the estimated fold change for the proposed method was on average 21.8\% smaller than $Limma$ IQR and 10.4\% smaller than $t$-test's IQR. This fold change comparison can be seen in {\bf Supplementary Figure~\ref{supp-fig:fc_boxplot}}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{In the label-free benchmark experiment, MSstatsPTM \todo{had a higher metric XXX than} the existing methods}

\todo{This section just describes the figure. Could you emphasize why we see these results? What are the reasons for differences in performance?} In this experiment all models incorrectly estimated the fold change of the modified spike-in peptides before adjusting for changes in global protein abundance. After adjustment, the spike-in peptides' fold change was generally in line with expectation in all methods, however the distribution of estimated fold changes was visibly wider (\figref{spikein_boxplot}). Of the three approaches, the proposed method showed the tightest distribution around the true log fold change. Comparing the IQR of the spike-in peptide's log fold change, the proposed method's IQR was 36.78\% smaller than $Limma$ and 32.98\% smaller than $t$-test's. While Limma and $t$-test generally estimated the correct fold change, the proposed method's estimation was more consistent across all modifications.

@@

@@

%\begin{table}[]
%\begin{tabular}{|c|c|c|}
%\hline
%Comparison & Unadj. true FC & Adj. true FC \\ \hline \hline
%mix1-mix2  & -1                           & 1                        \\ \hline
%mix1-mix3  & 0                           & -1                         \\ \hline
%mix1-mix4  & -1                          & 0                         \\ \hline
%mix2-mix3  & 1                          & -2                        \\ \hline
%mix2-mix4  & 0                           & -1                        \\ \hline
%mix3-mix4  & -1                           & 1                         \\ \hline
%\end{tabular}
%\end{table}

>>>>

@@

\medskip \noindent 
The basic $t$-test can be extended to account for the confounding of changes in PTM abundance and overall changes in protein abundance~\cite{Schwammle2015,THOMAS2020,Mertins:2013}. Typically this is done by first summarizing the feature intensities in each run, separately for modified and unmodified features, e.g. with the sum on the original scale, dividing the summary of the PTM abundance by the summary of the protein abundance and then taking a log. Specifically, for condition $i$ and biological replicate $j$, the approach calculates
\begin{eqnarray}
u_{ij} = \log \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right) - \log \left( \sum_{l=1}^{L} 2^{y_{ijl}} \right)
\label{eq:adj_estimation}
\end{eqnarray}
The approach then models these values as
\begin{eqnarray}
u_{ij} = (\mu^{\ast}_{i}-\mu_i) + \epsilon^{\prime}_{ij}, \text{where } \epsilon^{\prime}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{\prime 2}), i \in \{1, 2\}
\label{eq:ttest_model}
\end{eqnarray}
The model allows us to estimate $\hat{\Delta}_{adj}$ and its standard error.
Based on this model, we can test the more relevant null hypothesis in~\eqref{null_hyp}, by comparing the test statistic against the Student distribution with \todo{$df=2J-2$} \todo{this should be the same df as the previous section?} degrees of freedom in balanced designs.

Although effective, the approach is somewhat simplistic. It is not applicable to experimental designs with more complex sources of biological and technological variation, such as experiments with repeated measurements, experiments with multiple batches or experiments with multi-run multiplexing. Since \eqref{adj_estimation} performs the adjustment on the replicate level, the experiment must contain a matching number of replicates in both the modified and unmodified runs. Technological artifacts such as missing values further undermine the calculation of $u_{ij}$ in~\eqref{adj_estimation}. Finally, their is no self contained, straightforward implementation of the method, such as in the form of a coding package, requiring a more challenging manual application.

@@

$MSstats$~\cite{Choi:2014} and $MSstatsTMT$~\cite{Huang:2020} are a series of R/Bioconductor packages designed for protein significance analysis and statistical relative quantification of proteins and peptides in global, targeted and data-independent proteomics. Models in these packages takes as input a list of log$_2$-intensities $y_{ijk}$. For each protein, the feature log$_2$-intensities are first summarized into a single value per protein per run using Tukey's median polish~\cite{Tukey:1977}. These summarized values are then used as input to fit a flexible family of linear mixed-effects models \cite{McLean:1991, Faraway:2006, Bolker2009}. The specific model instances depend on the experimental design and data acquisition. For example, to model the unmodified protein in \figref{data-structure} $MSstats$ fits the following model
\begin{eqnarray}
& y_{ij} = \mu + Condition_i + Run_{j} + \epsilon_{ij}, & \\ 
& \text{where } \sum\limits_{i=1}^{J} Condition_i = 0,\ Run \mathop\sim^{iid} \mathcal{N}(0, \sigma^2_j), \epsilon_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^2) & \nonumber
\label{eq:msstats_model}
\end{eqnarray}
Parameters of the model are estimated using restricted maximum likelihood (REML)~\cite{Kenward}.
In turn, the parameters allow us to estimate the relevant quantity $\hat{\Delta}_{Protein}$ and its standard error. Where $\hat{\Delta}_{Protein}$ is estimated as

\begin{eqnarray}
\Delta_{Protein} = \mu_{i} - \mu_{i^{\prime}}
\end{eqnarray}

Similarly to $Limma$, $MSstatsTMT$ includes an optional Empirical Bayes moderation of the standard error~\cite{Huang:2020}. Using Empirical Bayes helps overcome cases where the number of biological replicates in each condition is small. Based on the model, the approach tests the null hypothesis in \eqref{conv_null_hyp}, by comparing the test statistic against the Student distribution.

The $MSstats$ framework has a number advantages over the methods above. 
First, unlike \todo{t-test (or ANOVA?) and $Limma$}, the family of linear mixed effects models is applicable to many complex experimental designs, including designs with multiple sources of variation, and unbalanced designs. 
Second, the approach is applicable to various data acquisition types, including label-free DDA and DIA, and experiments with TMT labeling, and is compatible with various data processing tools such as Skyline, Spectronaut, MaxQuant, Progenesis, Proteome Discoverer, and OpenMS. Third, the run-level summarization has been shown to be more robust to outliers and missing values as compared to feature averages~\cite{Choi:2014}. 
Finally, the custom $MSstats$ and $MSstatsTMT$ implementation accounts for potential data artifacts, and is numerically scalable and stable, and is available through both command line and a dedicated graphical user interface.

Unfortunately, MSstats estimates the difference in overall protein abundance and tests the null hypothesis in~\eqref{conv_null_hyp}. It does not account for the confounding between the changes in PTM abundance and the overall changes in protein abundance. In other words, it cannot directly test the null hypothesis in~\eqref{null_hyp}.

@@

We leverage the existing methods in $MSstats$ and $MSstatsTMT$ and extend them to remove confounding between changes in the PTM and the overall changes in protein abundance.  Each feature type is summarized separately using $MSstats$ or $MSstatsTMT$, resulting in run level values for the PTM, $\hat{y}_{ij}^{\ast}$, and unmodified peptide. The estimated summaries of the modified features are then used as the input to the model, such as in \eqref{msstats_model}, in $MSstats$ or $MSstatsTMT$. Similarly, the estimated summaries of the unmodified features are used as the input to the corresponding model in $MSstats$ or $MSstatsTMT$.

For the modified features, the model-based estimates in $MSstats$ and $MSstatsTMT$ include $\hat{\Delta}_{PTM}=\hat{\mu}^{\ast}_{i}-\hat{\mu}_{i^{\prime}}^{\ast}$, and its standard error $\widehat{SE}(\hat{\Delta}_{PTM})$. Similarly, for the unmodified features, the model-based estimates in $MSstats$ and $MSstatsTMT$ incluse $\hat{\Delta}_{protein}=\hat{\mu}_{i}-\hat{\mu}_{i^{\prime}}$, and its standard error $\widehat{SE}(\hat{\Delta}_{protein})$. Therefore, the adjusted difference $\hat{\Delta}_{adj}$ in~\eqref{null_hyp} can be estimated as
\begin{eqnarray}
\hat{\Delta}_{adj} = (\hat{\mu}^{\ast}_{i}-\hat{\mu}_{i}) - (\hat{\mu}_{i^{\prime}}^{\ast}-\hat{\mu}_{i^{\prime}}) = (\hat{\mu}^{\ast}_{i}-\hat{\mu}_{i^{\prime}}^{\ast}) - (\hat{\mu}_{i}-\hat{\mu}_{i^{\prime}}) = \hat{\Delta}_{PTM} - \hat{\Delta}_{protein} 
\label{eq:Delta_hat_adj}
\end{eqnarray}
Assuming that the unexplained by the model sources of variation in the modified features are independent from the corresponding sources of variation in the modified features, the standard error $\widehat{SE}(\hat{\Delta}_{adj})$ is obtained by combining the standard errors from the unmodified and modified model fits.
\begin{eqnarray}
\widehat{SE}(\hat{\Delta})_{adj} = \sqrt{ \widehat{SE}(\hat{\Delta}_{PTM})^{2} + \widehat{SE}(\hat{\Delta}_{protein})^{2}} 
\label{eq:prop_se}
\end{eqnarray}
The estimated standard error is larger than the standard errors associated with each distinct feature type, reflecting the increased uncertainty inherent to combining the uncertainty in two estimators. Finally, the degrees of freedom associated with~\eqref{prop_se} are obtained via the Satterthwaite approximation~\cite{kutner_etal_04a, satterthwaite:1946}
\begin{eqnarray}
\left( \widehat{SE}(\hat{\Delta}_{PTM})^{2} + \widehat{SE}(\hat{\Delta}_{protein})^{2} \right)^2 \bigg/
\left( \frac{\widehat{SE}(\hat{\Delta}_{PTM})^{4}}{\mathrm{df}(\widehat{SE}(\hat{\Delta}_{PTM}))} + \frac{\widehat{SE}(\hat{\Delta}_{protein})^{4}}{ \mathrm{df}(\widehat{SE}(\hat{\Delta}_{protein}))} \right)
\label{eq:prop_df}
\end{eqnarray}
To test the null hypothesis in~\eqref{null_hyp}, the test statistic $\hat{\Delta}_{adj}/\widehat{SE}(\Delta)_{adj}$ is compared with the Student distribution with the degrees of freedom in~\eqref{prop_df}. The p-values of the comparison are adjusted for multiple testing using the approach by Benjamin and Hochberg~\cite{Benjamini:1995}.

By leveraging the implementations in $MSstats$ and $MSstatsTMT$, the proposed approach is very versatile. It is applicable to a wide variety of experimental designs, including group comparison, paired designs, time course designs and unbalanced designs. It is applicable to label-free data acquisition strategies such as DDA and DIA and experiments with TMT labeling. The proposed approach is also applicable to experiments where the experimental designs for PTM profiling and global proteome profiling vary in properties such as number of biological replicates, data acquisition strategies and runs.

@@

The proposed methods are implemented in the open source R package $MSstatsPTM$. An overview of the steps of the package are illustrated in Figure \ref{fig:msstatsptm_design}. $MSstatsPTM$ includes converters for multiple spectral processing tools, including MaxQuant, Progenesis, and Spectronaut. Conversion is done separately for the modified peptide and global profiling runs. To identify the local modification site we require the tool to identify where in the sequence the modification occurred. If the global profiling information is not available, the package can still analyze the modified run, but will do so without adjusting for changes in unmodified protein abundance. This amounts to testing the null hypothesis in~\eqref{conv_null_hyp}; which amounts to running the methods seen in $MSstats$ on modified peptide level data.

The next step is peptide/protein summarization. Summarization is performed separately for the PTM and global protein in the same way as described above for $MSstats$. The package can optionally impute missing values using an Accelerated Failure Time (AFT) model~\cite{Wei:1992}. Missing value imputation should be performed with care when applied to PTM data. PTMs generally exhibit low feature counts and may be missing due to reasons other than low abundance. These issues can cause the missing value imputation to be unstable. Finally, a separate model is fit to both the modified and global profiling runs. This model is automatically adjusted depending on the experimental design and acquisition method. If the base model is not applicable, a simplified model is fit. After modeling, the modified model is adjusted for changes in unmodified protein abundance, using the methods described above.

We separately visualize the PTM and global protein side by side, allowing for easier interpretation and comparison across the datasets. In the modeling stage this is further separated into PTM, global protein, and adjusted PTM visualizations. This allows users to see the affect of the adjustment.

@@

The evaluations were based on the $MSstatsPTM$, and were performed without imputing missing values and without using Empirical Bayes moderation. $MSstatsPTM$ was compared with \todo{$t$-test?} and Limma. We did not compare to IsobarPTM because it is only applicable to TMT experiments, meaning it could not be applied to our ground truth data. All the evaluations were done at the adjusted p-value cutoff of $\alpha=.05$. More details are in {\bf Supplementary Sec. \ref{supp-sec:experiments}}.

@@

\figref{sim1_fdr} illustrated the high FDR observed on the simulated datasets when not accounting for the overall changes in protein abundance. After the adjustment, the FDR of all methods improved. In a clean simulation, recall and accuracy of $MSstatsPTM$ and $Limma$ performed similarly. As expected when not using EB moderation in MSstatsPTM, Limma performed slightly better in simulations with fewer biological replicates. In contrast, the performance of the two-sample $t$-test method lagged behind. This is due to the fact that the $t$-test only uses data within the groups of interest while ignoring the remainder of the data. In contrast, $MSstatsPTM$ and Limma leverage information across comparisons, resulting in improved power in simulations with more comparisons. Since this simulation is a group comparisons design that doesn't require many variance components, it is favorable to Limma and t-test

@@

\todo{This section just describes the figure. Could you emphasize why we see these results? What are the reasons for differences in performance?}

In simulations with missing values and few features, $MSstatsPTM$ outperformed $t$-test and $Limma$. Changes in unmodified protein level still needed to be accounted for to control the FDR. Once controlling for changes in the unmodified protein, the proposed method outperformed the other methods, as seen in~\figref{sim2_acc}. The proposed method calibrated model accuracy and recall rate well, even when the number of replicates were low. The proposed approach can handle low feature counts well due to using Tukey's Median Polish for summarization. The existing approaches summation approach require more observations to correctly summarize the intensities in the run, whereas TMP is more robust. \todo{more/better}

Additionally, when comparing the fold change estimation across all modified peptides, the proposed method showed a tighter distribution of estimated fold changes around the true fold change, as compared to $t$-test and $Limma$. Specifically, the inner quartile range (IQR) of the estimated fold change for the proposed method was on average 21.8\% smaller than $Limma$ IQR and 10.4\% smaller than $t$-test's IQR. This means that while the mean of the estimated fold changes was generally correct for all methods, the proposed approach correctly estimated the fold change more often across all PTMs. This fold change comparison can be seen in {\bf Supplementary Figure~\ref{supp-fig:fc_boxplot}}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{In the label-free benchmark experiment, MSstatsPTM had a higher sensitivity than the existing methods}

In this experiment all models incorrectly estimated the fold change of the modified spike-in peptides before adjusting for changes in global protein abundance. After adjustment, the spike-in peptides' fold change was generally in line with expectation in all methods, however the distribution of estimated fold changes was visibly wider (\figref{spikein_boxplot}). Of the three approaches, the proposed method showed the tightest distribution around the true log fold change. Comparing the IQR of the spike-in peptide's log fold change, the proposed method's IQR was 36.78\% smaller than $Limma$ and 32.98\% smaller than $t$-test's. While Limma and $t$-test generally estimated the correct fold change, the proposed method's estimation was more consistent across all modifications.

@@

This experiment introduced more variance components, which is challenging for the existing methods and therefore the difference in the results is more pronounced than in the simulations.

@@

%\begin{table}[]
%\begin{tabular}{|c|c|c|}
%\hline
%Comparison & Unadj. true FC & Adj. true FC \\ \hline \hline
%mix1-mix2  & 1                           & 1                        \\ \hline
%mix1-mix3  & 0                           & -1                         \\ \hline
%mix1-mix4  & 1                          & 0                         \\ \hline
%mix2-mix3  & -1                          & -2                        \\ \hline
%mix2-mix4  & 0                           & -1                        \\ \hline
%mix3-mix4  & 1                           & 1                         \\ \hline
%\end{tabular}
%\end{table}
